---
title: "Final project planning document"
author: "Sam Barbaro"
date: "2026-07-05"
output: html_document
---

## Final Project Planning Document

I will use this Beer Advocate data set from Kaggle, with 1.5 million ratings. It's located here:

<https://www.kaggle.com/datasets/thedevastator/1-5-million-beer-reviews-from-beer-advocate/data>

### Part 1

My beer purchases are typically based upon the name of the product and whether it purports to be cake-flavored (the second thing is almost always a lie, but hope springs eternal). Because you don't really know much about a drink when you're buying it (it's not like a book where you can read the first few pages or a movie where you can watch the trailer or read a description), purchases are made based on limited information. For these reasons, I would like to take the Beer Advocate data set, implement ALS, and optimize it. I want to predict whether someone would be interested in trying a beer. I also want to try deploying the same code in Azure, to see if using the dedicated RAM speeds up the process and creates efficiencies. For example, below, I would have had a dataframe/matrix that was 56k rows wide, and I had to pare down by deleting beers that had ten or fewer ratings. The entire data set may be easier to run in Azure.

**The steps:**

- Clean the data

- Explore the data (how many ratings are there? What is the average rating? Which beers are the most-rated? What is the average number of ratings? Is there a most popular type of beer? Are there products with only one or two ratings?)

- Create a binary matrix

- Implement ALS

- Optimize parameters, including lambda, n_factors, and n_iterations

- Implement ALS again

- Visualize results

- Run optimized ALS in Azure and document results/times

### Part 2: incorporating images

If there's time, I will attempt this in a more limited capacity than part 1.

I believe labels are a big factor in drink purchases, especially with beer, where costs are roughly similar across the board. However, unlike movie posters, it's hard to find images for every beer: some of these products may no longer exist, Beer Advocate uses a placeholder image for many products, and photos are not in a centralized place, like IMDB. I may pull a subset of the most popular beers' images and attempt to incorporate image processing into collaborative filtering to see if features like color or shape play a role in beer choices. This will be tested locally and will involve a subset of the data.

### Preliminary data cleaning

Below, there's some code making sure I can transform this file (which has about 56k products) into wide format.

```{r, warning=FALSE}
library(tidyverse)
library(recommenderlab)

beer_ratings <- read.csv(
  "C:/Users/Sam Barbaro/Desktop/Data 612/beer_reviews.csv", 
  na.strings = c("", "NA", "null", "NULL"),  header = TRUE)

```

Select relevant columns

```{r}
beer_users <- beer_ratings |>
  select(review_profilename, beer_name, review_overall)

#1.6 million-ish rows 
count(beer_users)

#how wide will this matrix be?
distinct_beers <- beer_users |> distinct(beer_name
                                        )

distinct_users <- beer_users |> distinct(review_profilename
                                        )


```

There are 56k beers, meaning the matrix will be 56,857 rows wide. There are 33,388 users. It's possible there are some beers that have only been reviewed a few times.

```{r}
#how many beers have fewer than 5 reviews? 

beer_counts <- beer_users |>
    group_by(beer_name) |>
    summarise(n_reviews = n()) |>
    arrange((n_reviews))

beer_less_than5 <- beer_counts |>
  filter(n_reviews < 11)

#about 38k beers have fewer than 5 reviews
count(beer_less_than5)

```

I was unable to make a grid that was 56k wide, so I will eliminate beers with 10 or fewer reviews.

```{r}
beer_userids <- beer_users |>
  mutate(user_id = as.integer(factor(review_profilename))) |>
  select(user_id, beer_name, review_overall)

#anti-join (this should  happen before the previous step in 
#the final version)

beer_reduced <- anti_join(beer_userids, beer_less_than5, by = "beer_name")


```

Create the wide table before converting to a real ratings matrix

```{r}
beer_wide <- pivot_wider(
  data = beer_reduced,
  id_cols = user_id,
  names_from = beer_name,
  values_from = review_overall,
  #in case users reviewed a beer more than once (reviewers can)
  values_fn = first 
)

beer_wide <- beer_wide |> arrange(user_id)
```
