---
title: "Project 1 Data 612"
format: html
editor: visual
---

## Project 1

This system recommends rom-com novels to readers.

### Loading the data

```{r}
library(tidyverse)


romcom_books <- read.csv("https://raw.githubusercontent.com/samanthabarbaro/data612_recommender_systems/refs/heads/main/proj_1_dataset.csv", na.strings = c("", "NA", "null", "NULL"),  header = TRUE)
```

## Splitting the data

Split the data into an 80/20 training and test set: I've pivoted longer so it's easier to split up different ratings from the same users. With this format, using a random sample of the data set, some users are represented in the training data but not the test data.

```{r}
pivoted_books <- romcom_books |>
  pivot_longer(
  cols = Carry.On:Attachments,
        names_to = "book",
        values_to = "rating"
    )

pivoted_books
```

### Create training and test data frames

```{r}
set.seed(1122)

#remove NAs and rename the user column

pivoted_books_2 <- pivoted_books |>
  filter(!is.na(rating)) |>
  rename(user = X)
  

#create a training and test set

train_index <- sample(nrow(pivoted_books_2),
                      0.8 * nrow(pivoted_books_2))

train_books <- pivoted_books_2[train_index, ]

test_books <- pivoted_books_2[-train_index, ]


```

### Calculating the raw average and RMSE for training and test data

```{r}
#mean overall rating (raw average)

raw_average <- train_books |> 
  mutate(mean_rating = round(mean(rating),2))

error <- raw_average |>
  #calculate the difference (error)
  mutate(difference = (rating - mean_rating)) |>
  #square the errors
  mutate (square = (difference * difference))

error

#number of rows in the training set = 62
count(train_books)

#using 62 is not dynamic (does not account for adjusted sizes of training sets)
#which is fine for this one instance, but could be substituted with count(train_books)

#calculate rmse
rmse_training <- error |> 
  #sum the squared errors
  summarise(total_sq = sum(square)) |>
  #take the mean and then the sq root
  mutate(rmse = (sqrt(total_sq/62))) 

#the rmse for the training set is 1.16
rmse_training

#number of rows in the test set is 16
count(test_books)

#calculate RMSE for test set
rmse_test <- test_books |> 
  mutate(mean_rating = 3.23) |>
  mutate(difference = (rating - mean_rating)) |>
  mutate(square = (difference*difference))

rmse_test <- rmse_test |>
    summarise(total_sq = sum(square)) |>
  #take the mean and then the sq root
  mutate(rmse = (sqrt(total_sq/16))) 

#RMSE for the test set is 1.295
rmse_test

```

### Calculate the bias for each user and each item

The bias for each user is the mean of their ratings - the mean across all books (how much higher or lower they tend to rate something).

The bias for each item is how much higher or lower the average rating is than the over all average rating.

User bias = user average - raw average

Item/book bias = movie average - raw average

```{r}
#mean for each book, user, and overall
item_user_mean <- train_books |> 
  #mean for each book
  group_by(book) |> 
  mutate(book_mean = round(mean(rating),2)) |> 
  ungroup() |>
  #mean for each user
  group_by(user) |> 
  mutate(user_mean = round(mean(rating),2)) |> 
  ungroup() |>
  mutate(overall_mean = 3.23)

#calculate the biases

# user bias = user mean - overall mean (across all books)

#item bias = item mean - overall mean (across all books)

#creating a key that I will join to the original pivoted data set, which includes all
#item/user combos 
user_biases <- item_user_mean |> 
  mutate(user_bias = (user_mean - overall_mean)) |>
  distinct(user, user_bias)

item_biases <- item_user_mean |> 
  mutate(item_bias = (book_mean - overall_mean)) |>
  distinct(book, item_bias)

```

### Calculating baseline predictors

Finding the baseline predictors for every user-item combination.

Baseline predictor = raw average + user bias + movie bias

```{r}
book_key <- pivoted_books |> 
  rename(user = X) |>
  mutate(overall_mean = 3.23) |>
  left_join(user_biases, by = "user") |>
  left_join(item_biases, by = "book")

predictions <- book_key |>
  mutate(baseline_predictor = (overall_mean + user_bias + item_bias)) |>
  #if it's larger than 5, then 5
  mutate(baseline_predictor = if_else(baseline_predictor > 5, 5, baseline_predictor))

predictions
```

Select from the pivoted longer data set (before removing NAs) just the user and the movie column, then join with the movie average rating:

```{r}
train_predictions <- train_books |>
  left_join(predictions, by = c("user", "book"))

train_predictions

test_predictions <- test_books |> 
   left_join(predictions, by = c("user", "book"))

test_predictions
```

Calculating RMSE for the baseline predictors for both the training and test data:

```{r}
#calculate the difference and square the difference

train_predictions <- train_predictions |>
  mutate(difference = rating.x - baseline_predictor) |>
  mutate(square = (difference*difference))

rmse_training_baseline <- train_predictions |> 
  #sum the squared errors
  summarise(total_sq = sum(square)) |>
  #take the mean and then the sq root
  mutate(rmse = (sqrt(total_sq/62))) 

#the rmse for the training set is .55
rmse_training_baseline
```

RMSE for test set:

```{r}
test_predictions <- test_predictions |>
  mutate(difference = rating.x - baseline_predictor) |>
  mutate(square = (difference*difference))

rmse_test_baseline <- test_predictions |> 
  #sum the squared errors
  summarise(total_sq = sum(square)) |>
  #take the mean and then the sq root
  mutate(rmse = (sqrt(total_sq/16))) 

#the rmse for the test set is much higher at .935
rmse_test_baseline
```

### Summary

The RMSE was highest when using the raw average, which doesn't account for differences across users or media at 1.295 on the test data and 1.16 on the training data.

Results were slightly better when accounting for user and book bias. When I calculated the predicted rating for the test set, the RMSE was only .55. However, it was much higher for the test set at .935 (almost a full point). When checking the calculations, one large difference (a low rating of a generally highly rated book by user 15, a high rater) may have introduced a lot of error.

The percent improvement on test data, when switching from a raw average method to a baseline predictor method, was about 28% on the test data and about 52% on the training data.

```{r}
test_improvment <- (1 - .935 /1.295)
training_improvement <- (1 - .555/1.16)
test_improvment
training_improvement
```
